Integrated multi-omics analysis reveals Lactobacillus anti-inflammatory process in vaginal tissue

### 2021-02-20 reports/rmarkdown/03_integrated_datasets.Rmd
adds picrust and RNAseq to manuscript

NESse_list <- list()

@@

  allNES <- setNames(rep(NA,length(allpathways)),allpathways)
  NESes <- lapply(enrichments,function(x) {
    temp <- allNES
    temp[x$pathway] <- x$NES
    return(temp)
  })
  NESes <- t(as.data.frame(NESes))
  # NESes[is.na(NESes)] <- 0
  # NESes[ pvalues < -log10(0.1) ] <- 0

  NESse_list[[i]] <- NESes

@@

write.csv2(NESse_list,"../../results/Hallmark_NESse_list.csv")

@@

  allNES <- setNames(rep(NA,length(allpathways)),allpathways)
  NESes <- lapply(enrichments,function(x) {
    temp <- allNES
    temp[x$pathway] <- x$NES
    return(temp)
  })
  NESes <- t(as.data.frame(NESes))
  # NESes[is.na(NESes)] <- 0
  # NESes[ pvalues < -log10(0.1) ] <- 0

  NESse_list[[i]] <- NESes

write.csv2(NESse_list,"../../results/Kegg_NESse_list.csv")

>>>>

NESse_list <- list()
pvalues_list <- list()

@@

  allNES <- setNames(rep(NA,length(allpathways)),allpathways)
  NESes <- lapply(enrichments,function(x) {
    temp <- allNES
    temp[x$pathway] <- x$NES
    return(temp)
  })
  NESes <- t(as.data.frame(NESes))
  NESes[is.na(NESes)] <- 0
  # NESes[ pvalues < -log10(0.1) ] <- 0

  NESse_list[[i]] <- NESes
  pvalues_list[[i]] <- pvalues

@@

write.csv2(NESse_list,"../../results/Hallmark_NESse_list.csv")
write.csv2(pvalues_list,"../../results/Hallmark_pvalues_list.csv")

@@

  allNES <- setNames(rep(NA,length(allpathways)),allpathways)
  NESes <- lapply(enrichments,function(x) {
    temp <- allNES
    temp[x$pathway] <- x$NES
    return(temp)
  })
  NESes <- t(as.data.frame(NESes))
  NESes[is.na(NESes)] <- 0
  # NESes[ pvalues < -log10(0.1) ] <- 0

  NESse_list[[i]] <- NESes
  pvalues_list[[i]] <- pvalues
}

write.csv2(NESse_list,"../../results/Kegg_NESse_list.csv")
write.csv2(pvalues_list,"../../results/Kegg_pvalues_list.csv")

### 2021-02-20 reports/rmarkdown/manuscript/manuscript_template.Rmd
adds picrust and RNAseq to manuscript

# I usually load my libraries up front to keep things organized
library(bookdown)
library(knitr)
library(kableExtra)
library(ggplot2)
library(igraph)
library(dplyr)
library(stringr)
library(rafalib)
library(scales)
# remotes::install_github("czarnewski/niceRplots",force=T)
library(niceRplots)

@@

picrust_data <- read.table("../../../data/pred_metagenome_unstrat.tsv",header = T,row.names = 1)
picrust_dge <- read.csv("../../../data/Picrust2_top_functions.csv",header = T,row.names = 1)
```

@@

metadata$group_annotation <- group_annotation
```

@@

par(mar=c(0,0,0,5))
plot( g , vertex.label.cex=0.000001 , vertex.color = pal[factor(group_annotation)] ,
      edge.width=  ( E(g)$weight / max(E(g)$weight)) ,
      vertex.size=10,main="", 
      edge.color=colorRampPalette(c("grey95","black"))(90) [ round( E(g)$weight / max(E(g)$weight) * 88 )+1 ] )
legend("topright",title.adj = 0,
       legend = levels(factor(group_annotation)),xjust = 0,yjust = 1,
       bty = "n",pch = 21,pt.bg = pal,pt.cex = 1,xpd=T)

@@

```{r, echo = FALSE, eval = TRUE}
# Define Layout and set labels
layout(matrix(c(2 ,3,4,5,      6,7,17,17,          19,19,19,19,
                1,1,1,1,      1,8,17,17,          20,20,20,20,
                1,1,1,1,      1,9,18,18,         21,21,24,24,
                1,1,1,1,      1,10,18,18,         22,22,24,24,
                16,15,14,13,  12,11,23,23,         23,23,24,24 ) , ncol= 12,byrow = T),
       widths = 1,heights = c(1.2,1.2,1.2,1.2,1.2))
# layout.show(12)
figlabels <- letters

#########################
# BACTERIAL COMMUNITIES #
#########################
gB <- graph_from_adjacency_matrix(as.matrix(SNN_bacteria), mode = "undirected",diag = F,weighted = T)
set.seed(1)
lB <- layout_with_fr(gB,niter=3000,start.temp=30)

par(mar=c(0,0,0,5))
plot( gB , vertex.label.cex=0.000001 , vertex.color = pal[factor(bac_communities)] ,
      edge.width=  ( E(gB)$weight / max(E(gB)$weight)) ,
      vertex.size=10,main="", 
      edge.color=colorRampPalette(c("grey95","black"))(90) [ round( E(gB)$weight / max(E(gB)$weight) * 88 )+1 ] )
legend("topright",title.adj = 0,
       legend = levels(factor(bac_communities)),xjust = 0,yjust = 1,
       bty = "n",pch = 21,pt.bg = pal,pt.cex = 1,xpd=T)

@@

par(mar=c(1,1,1,1))
for(j in sort(unique(bac_communities)) ){
  empty_plot()
  text( 0,1,labels = paste0(rownames(SNN_bacteria)[ bac_communities == j ],collapse = "\n"),adj=c(0,1),cex=.5)
  title(main=paste0("community",j),cex.main=.7)
}
for(i in 1:(18-length(unique(bac_communities)))){empty_plot()}

@@

par(mar=c(1,7,2,3))
x <- taxa[match(rownames(SNN_bacteria),rownames(taxa)),2]
y <- taxa[match(rownames(SNN_bacteria),rownames(taxa)),3]
# plot_sankey( data.frame( x[y!="NA"] , y[y!="NA"]), color_by = 1,
#        plot_labels = T, plot_weights = F,order_1_by = "total",
#        order_2_by = "total",use_w1 = T,use_w2 = T,gapv = .02,gap2v = 0,smoothing = .1,
#        pal=hue_pal()(length(unique(x))))

@@

plot_sankey( data.frame( x[y!="NA"] , bac_communities[y!="NA"]), color_by = 1,
     plot_labels = T, plot_weights = F,order_1_by = "total",
     order_2_by = "total",use_w1 = T,use_w2 = T,gapv = .02,gap2v = 0,smoothing = .1,
     pal=hue_pal()(length(unique(x))))

@@

##############################################
# BACTERIAL COMMUNITY DIFFERENTIAL ABUNDANCE #
##############################################
par(mar=c(1,7,1,3))

module_means_per_dataset <- lapply(datasets_all_samples[-1],function(x){
  return( rowsum( x[rownames(SNN_bacteria),] , group = bac_communities ) / c(table(bac_communities)) )
})

for( j in c("ASV_tissue_V3_normalized_batch_corrected",
        "ASV_CVL_V3_normalized_batch_corrected",
        "ASV_CVL_V2_normalized_NOT_batch_corrected")){
  temp_group <- as.character(group_annotation)
  temp_group[ is.na(temp_group) ] <- "NA"
  temp_group[ temp_group == "" ] <- "NA"
  temp_group <- factor(temp_group)

  barlist(data=module_means_per_dataset[[j]],
          genes=rownames(module_means_per_dataset[[j]]),col = pal,
          clustering=temp_group, main=paste0("\n", sub("_norm.*","",j)))
}
add_letter(figlabels[1]); figlabels <- figlabels[-1]

empty_plot()

#####################
# BACTERIAL PICRUST #
#####################
par(mar=c(1,7,1,3))
# 
# library(keggorthology)
# data("KOgraph")
# annot <- as.data.frame(unlist(nodeData(KOgraph,attr="tag")))
# colnames(annot)[1] <- "KEGGID"
# annot$pathway <- rownames(annot)
# annot$KEGGID <- paste0("K",annot$KEGGID)
# 
# library(KEGGREST)

download.file("https://raw.githubusercontent.com/picrust/picrust2/master/picrust2/default_files/description_mapfiles/ko_info.tsv.gz","../../../data/KEGG_pathways_to_KO.tsv")
download.file("https://raw.githubusercontent.com/picrust/picrust2/master/picrust2/default_files/pathway_mapfiles/KEGG_pathways_to_KO.tsv","../../../data/KEGG_pathways_to_KO.tsv")
download.file("https://raw.githubusercontent.com/picrust/picrust2/master/picrust2/default_files/description_mapfiles/KEGG_pathways_info.tsv.gz","../../../data/KEGG_pathways_info.tsv.gz")
system("gunzip ../../../data/*.gz")

kegg_pathway_annot <- fgsea::gmtPathways('../../../data/KEGG_pathways_to_KO.tsv')

kegg_names <- read.delim('../../../data/KEGG_pathways_info.tsv',header = F)

ko_names <- readLines('../../../data/ko_info.tsv')
ko_names <- strsplit(ko_names,"\t")

path_trend <- t(sapply(kegg_pathway_annot,function(x){
  x <- rownames(picrust_data)[rownames(picrust_data) %in% x]
  if(length(x)>=2){
    return(colSums(picrust_data [ x , ]))
  } else {
    return(rep(0, ncol(picrust_data)))
  }
}))

path_trend <- path_trend[rowSums(path_trend)!=0 , ]
rownames(path_trend) <- kegg_names[ match(rownames(path_trend),kegg_names[,1]),2]

library(edgeR)
design <- model.matrix(~  group_annotation, data=metadata)

y <- DGEList(counts=floor(path_trend), remove.zeros = T)
y <- calcNormFactors(y,method = "TMM")
y <- estimateGLMCommonDisp(y,design)
y <- estimateGLMTagwiseDisp(y,design)
fit <- glmFit(y, design)

lrt <- glmLRT(fit,coef=1:8)
top <- topTags(lrt,adjust.method = "BH",n = "all",sort.by = "p.value")[[1]]
colnames(top) <- sub("group_annotation","",colnames(top))
top <- cbind(logFC.Gard=0,top)

norm_path_trend <- log2( t( t(path_trend) / colSums(path_trend) ) * 1000 + 1 )
norm_path_trend <- norm_path_trend[top_dge, ]
norm_path_trend <- t(apply(norm_path_trend,1,function(x){scale(x,T,T)}))
norm_path_trend[norm_path_trend > 5] <- 5
colnames(norm_path_trend) <- colnames(path_trend)
sample_use <- colnames(norm_path_trend)

mypar(2,4,mar=c(1,.2,2,.2))
h <- hclust( as.dist( (1- cor(t(norm_path_trend)))/2 ), method = "ward.D2")
plot( rev(as.dendrogram(h)) , horiz = T, leaflab = "none",xaxs="i",yaxs="i",axes=F)
abline(v=1,xpd=F,col="red",lty=2)
gene_module <- cutree(h, h = 1)

points( rep(0,length(gene_module)),
        seq(length(gene_module),1,length.out = length(gene_module)),
        col=taxa_pal[factor(gene_module[h$order])],
        pch=16,cex=.5)

image( t(norm_path_trend[h$order,][nrow(norm_path_trend):1,order(group_annotation[sample_use])]),
       col = colorRampPalette(c("navy","navy","grey95","firebrick4","firebrick4") )(99),
       breaks = seq(-5,5,length.out = 100),axes=F)
points( seq(0,1,length.out = length(sample_use)),rep(par("usr")[4],length(sample_use)),
        col=pal[group_annotation[sample_use][ order(group_annotation[sample_use])]],
        pch=16)

par(mar=c(4,4,2,2))
set.seed(1)
UMAP_picrust <- uwot::umap(t(norm_path_trend),n_neighbors = 20,
                 metric = "correlation",min_dist = .2,spread = .4,
                 repulsion_strength = .4,negative_sample_rate = 3)
plot(UMAP_picrust,bg=pal[group_annotation[sample_use]],main="UMAP on top functional profile",
     pch=21,xlab="UMAP1",ylab="UMAP2",frame=F,axes=F)
```

@@

mypar(2,4,mar=c(1,.2,2,.2))
h <- hclust( as.dist( (1- cor(t(top_TRX_counts)))/2 ), method = "ward.D2")
plot( rev(as.dendrogram(h)) , horiz = T, leaflab = "none",xaxs="i",yaxs="i",axes=F)
abline(v=2,xpd=F,col="red",lty=2)
gene_module <- cutree(h, h = 2)

points( rep(0,length(gene_module)),
        seq(length(gene_module),1,length.out = length(gene_module)),
        col=taxa_pal[factor(gene_module[h$order])],
        pch=16,cex=.5)

image( t(top_TRX_counts[h$order,][nrow(top_TRX_counts):1,order(group_annotation[sample_use])]),
       col = colorRampPalette(c("navy","navy","grey95","firebrick4","firebrick4") )(99),
       breaks = seq(-5,5,length.out = 100),axes=F)
points( seq(0,1,length.out = length(sample_use)),rep(par("usr")[4],length(sample_use)),
        col=pal[group_annotation[sample_use][ order(group_annotation[sample_use])]],
        pch=16)

set.seed(1)
UMAP_TRX <- uwot::umap(t(top_TRX_counts),n_neighbors = 10,
                 metric = "correlation",min_dist = 0.1,spread = 5,
                 negative_sample_rate = 10)
plot(UMAP_TRX,bg=pal[group_annotation[sample_use]],main="UMAP on top DGE RNAseq",
     pch=21,xlab="UMAP1",ylab="UMAP2",frame=F,axes=F)
```

@@

empty_plot()
legend(x = par("usr")[c(1)],
         y = par("usr")[c(4)],title.adj = 0,
         legend = rownames(temp),xjust = 0,yjust = 1,
         bty = "n",pch = 22,pt.cex = 2,cex = 1,xpd=T,
          title="top abundant bacteria",
         pt.bg = taxa_pal)

@@

>>>>

# I usually load my libraries up front to keep things organized
library(bookdown)
library(knitr)
library(kableExtra)
library(ggplot2)
library(igraph)
library(dplyr)
library(stringr)
library(rafalib)
library(scales)
library(dendextend)
library(edgeR)
remotes::install_github("czarnewski/niceRplots",force=T)
library(niceRplots)

@@

picrust_data <- read.table("../../../data/pred_metagenome_unstrat.tsv",header = T,row.names = 1)
```

@@

metadata$group_annotation <- group_annotation
write.csv2(metadata,"../../../results/metadata_integration.csv",row.names = T)

@@

par(mar=c(0,0,0,6))
plot( g , vertex.label.cex=0.000001 , vertex.color = pal[factor(group_annotation)] ,
      edge.width=  ( E(g)$weight / max(E(g)$weight)) ,
      vertex.size=10,main="", layout=l,
      edge.color=colorRampPalette(c("grey95","black"))(90) [ round( E(g)$weight / max(E(g)$weight) * 88 )+1 ] )
legend(par("usr")[2],par("usr")[4],title.adj = 0,
       legend = levels(factor(group_annotation)),xjust = 0,yjust = 1,
       bty = "n",pch = 21,pt.bg = pal,pt.cex = 1,xpd=T)

@@

```{r, echo = FALSE, eval = TRUE,fig.asp=.6}
# Define Layout and set labels
layout(matrix(c(2 ,3,4,5,      6,7,17,17,          17,17,17,17,
                1,1,1,1,      1,8,17,17,          17,17,17,17,
                1,1,1,1,      1,9,18,18,         18,18,18,18,
                1,1,1,1,      1,10,19,19,         19,19,19,19,
                16,15,14,13,  12,11,20,20,         20,20,20,20 ) , ncol= 12,byrow = T),widths = 1,heights = c(1.2,1.2,1.2,1.2,1.2))
# layout.show(12)
figlabels <- letters

#########################
# BACTERIAL COMMUNITIES #
#########################
gB <- graph_from_adjacency_matrix(as.matrix(SNN_bacteria), mode = "undirected",diag = F,weighted = T)
# set.seed(1)
lB <- layout_nicely(gB,niter=3000,start.temp=30)

par(mar=c(0,0,0,10))
plot( gB , vertex.label.cex=0.000001 , vertex.color = pal[factor(bac_communities)] ,
      edge.width=  ( E(gB)$weight / max(E(gB)$weight)) ,
      vertex.size=10,main="", 
      edge.color=colorRampPalette(c("grey95","black"))(90) [ round( E(gB)$weight / max(E(gB)$weight) * 88 )+1 ] ,layout=lB)
legend(par("usr")[2],par("usr")[4],title.adj = 0,
       legend = levels(factor(bac_communities)),xjust = 0,yjust = 1,
       bty = "n",pch = 21,pt.bg = pal,pt.cex = 1,xpd=T)

@@

par(mar=c(1,1,1,1))
for(j in sort(unique(bac_communities)) ){
  empty_plot()
  text( 0,1,labels = paste0(rownames(SNN_bacteria)[ bac_communities == j ],collapse = "\n"),adj=c(0,1),cex=.5)
  title(main=paste0("community",j),cex.main=.7)
}

@@

par(mar=c(1,10,1,15))
x <- taxa[match(rownames(SNN_bacteria),rownames(taxa)),2]
y <- taxa[match(rownames(SNN_bacteria),rownames(taxa)),3]
# plot_sankey( data.frame( x[y!="NA"] , y[y!="NA"]), color_by = 1,
#        plot_labels = T, plot_weights = F,order_1_by = "total",
#        order_2_by = "total",use_w1 = T,use_w2 = T,gapv = .02,gap2v = 0,smoothing = .1,
#        pal=hue_pal()(length(unique(x))))

@@

plot_sankey( data.frame( x[y!="NA"] , bac_communities[y!="NA"]), color_by = 1,
     plot_labels = T, plot_weights = F,order_1_by = "total",
     order_2_by = "total",use_w1 = T,use_w2 = T,gapv = .02,gap2v = 0,smoothing = .1,
     pal=hue_pal()(length(unique(x))),xaxs="i",yaxs="i")

@@

##############################################
# BACTERIAL COMMUNITY DIFFERENTIAL ABUNDANCE #
##############################################
module_means_per_dataset <- lapply(datasets_all_samples[-1],function(x){
  return( rowsum( x[rownames(SNN_bacteria),] , group = bac_communities ) / c(table(bac_communities)) )
})
# 
par(mar=c(1,10,1,6))
# for( j in c("ASV_tissue_V3_normalized_batch_corrected",
#         "ASV_CVL_V3_normalized_batch_corrected",
#         "ASV_CVL_V2_normalized_NOT_batch_corrected")){
#   temp_group <- as.character(group_annotation)
#   temp_group[ is.na(temp_group) ] <- "NA"
#   temp_group[ temp_group == "" ] <- "NA"
#   temp_group <- factor(temp_group)
#   
#   plot_dots(data=module_means_per_dataset[[j]],
#           genes=rownames(module_means_per_dataset[[j]]),col = pal,
#           clustering=temp_group, main=paste0("\n", sub("_norm.*","",j)))
# }
# add_letter(figlabels[1]); figlabels <- figlabels[-1]

plot_dots_difference(data = module_means_per_dataset[[2]],
                     data2 = module_means_per_dataset[[3]],
                     clustering = as.character(group_annotation),
                     clustering2 = as.character(group_annotation),
                     genes = rownames(module_means_per_dataset[[1]]),
                     main = "CVL3 vs CVL2",cex.row = .8)

plot_dots_difference(data = module_means_per_dataset[[2]],
                     data2 = module_means_per_dataset[[1]],
                     clustering = as.character(group_annotation),
                     clustering2 = as.character(group_annotation),
                     genes = rownames(module_means_per_dataset[[1]]),
                     main = "CVL3 vs tissue_V3",cex.row = .8)
# mypar(mar=c(2,2,2,6))
# plot(c(-2,10),c(-2,10),xaxs="i",yaxs="i")
# rrr <- rev(as.raster(colorRampPalette(c("blue","navy","grey95","firebrick","red"))(99)))
# rasterImage(as.raster(rrr),
#             xleft = par("usr")[2]+diff(par("usr")[1:2])/40,
#             xright = par("usr")[2]+diff(par("usr")[1:2])/20,
#             ybottom = par("usr")[4]-diff(par("usr")[1:2])/4,
#             ytop = par("usr")[4], 
#             interpolate = FALSE )
# text( c(par("usr")[2]+diff(par("usr")[1:2])/20,
#         par("usr")[2]+diff(par("usr")[1:2])/20) ,
#       c(par("usr")[4]-diff(par("usr")[3:4])/4,
#         par("usr")[4]),labels = c("min","max"),pos=4 )
# 
# 
# text( c(par("usr")[2]+diff(par("usr")[1:2])/20,
#         par("usr")[2]+diff(par("usr")[1:2])/20,
#         par("usr")[2]+diff(par("usr")[1:2])/20) ,
#       c(par("usr")[4]-diff(par("usr")[1:2])/4,
#         par("usr")[4]-diff(par("usr")[1:2])/8,
#         par("usr")[4]),labels = c("-1","0","+1"),pos=4 )
# text( c(par("usr")[2]+diff(par("usr")[1:2])/40) ,xpd=T,
#       c(par("usr")[4]+diff(par("usr")[3:4])/40),labels = c("Z-score"),adj=0 )

#####################
# BACTERIAL PICRUST #
#####################
picrust_data <- round(picrust_data[ rowSums(picrust_data>5)>=10, ])
sample_use <- colnames(picrust_data)

ko_names <- readLines('../../../data/ko_info.tsv')
ko_names <- strsplit(ko_names,"\t")
ko_names <- setNames(unlist(lapply(ko_names,function(x){x[2]})), unlist(lapply(ko_names,function(x){x[1]})))
ko_names <- sub(" [[].*","",ko_names)
ko_names <- sub(".*; ","",ko_names)

named_picrust_data <- rowsum( picrust_data , group = ko_names[ rownames(picrust_data) ] )
named_picrust_data[1:20,1:3]
dim(named_picrust_data)
named_picrust_data <- round(named_picrust_data[ rowSums(named_picrust_data>5)>=5, ])

design <- model.matrix(~group_annotation, data=metadata[sample_use,])

y <- DGEList(counts=named_picrust_data[,sample_use])
y <- calcNormFactors(y,method = "TMM")
y <- estimateGLMCommonDisp(y)
y <- estimateGLMTagwiseDisp(y)
fit <- glmFit(y, design)

top_dge <- (top$FDR < 1e-5) #& (rowSums( abs(top[,grep("logFC",colnames(top))]) >= log2(1.5) ) >= 1)
top_dge <- rownames(top)[ top_dge ]

top_picrust_data <- cpm(y)[top_dge, ]
top_picrust_data <- t(apply(top_picrust_data,1,function(x){scale(x,T,T)}))
top_picrust_data[top_picrust_data > 5] <- 5
colnames(top_picrust_data) <- colnames(picrust_data)

layout(matrix(c(1,2,3,4,
              1,2,3,4),
              nrow = 2,ncol = 4,byrow = T),widths = c(1,2,.5,6))
par(mar=c(4,.1,2,.1))

h <- hclust( as.dist( (1- cor(t(top_picrust_data)))/2 ), method = "ward.D2")
plot( rev(as.dendrogram(h)) ,xlim=c(max(h$height),-1), horiz = T, leaflab = "none",xaxs="i",yaxs="i",axes=T)
cutoff <- 4
abline(v=cutoff,xpd=F,col="red",lty=2)
gene_module <- cutree(h, h = cutoff)

points( rep(-.5,length(gene_module)),
        seq(length(gene_module),1,length.out = length(gene_module)),
        col=taxa_pal[factor(gene_module[h$order])],
        pch=16,cex=1,xpd=T)

image( t(top_picrust_data[h$order,][nrow(top_picrust_data):1,order(group_annotation[sample_use])]),
       col = colorRampPalette(c("navy","navy","grey95","firebrick4","firebrick4") )(99),
       breaks = seq(-5,5,length.out = 100),axes=F)
points( seq(0,1,length.out = length(sample_use)),rep(par("usr")[4],length(sample_use)),
        col=pal[group_annotation[sample_use][ order(group_annotation[sample_use])]],
        pch=16)

source("../enrichment_function.R")
gmt_list <- fgsea::gmtPathways("../../../data/KEGG_pathways_to_KO.tsv")
kegg_names <- read.delim('../../../data/KEGG_pathways_info.tsv',header = F)
names(gmt_list) <- kegg_names [ match ( names(gmt_list), as.character(kegg_names[,1]) ) , 2 ]
gmt_list <- lapply(gmt_list , function(x){
  na.omit(ko_names[x])
})
res_list <- lapply( unique( gene_module[h$order] ),function(x){
  temp <- compute_enrichment(genes = names(gene_module)[gene_module==x],
                            gmt_list = gmt_list,
                            min_terms_pathway = 10,
                            max_terms_pathway = 300,
                            min_overlap = 3,
                            sort_by_pvalue = T)
  temp <- temp[!grepl("REGULATION",rownames(temp)),]
  return(temp)
} )
names(res_list) <- unique( gene_module[h$order] )

pvalues <- unlist(lapply(res_list,function(x){ -log10(as.numeric(x$pvalue)) [1:5] }))
pvalues[is.na(pvalues)] <- 0
terms <- unlist(lapply(res_list,function(x){ rownames(x) [1:5] }))
terms[is.na(terms)] <- ""
genes <- unlist(lapply(res_list,function(x){ x$genes [1:5] }))
genes[is.na(genes)] <- ""

pvalues <- setNames(pvalues,paste0(terms," (",genes,")"))
module_color <- unlist(lapply(names(res_list),function(x){ rep(x,5) }))

temp <- factor( rev(gene_module), levels = unique( gene_module[h$order] ))
plot_sankey( data.frame(temp,temp), pal = taxa_pal[as.numeric(levels(temp))] , use_w2 = F , plot_labels = T ,gapv = .01,gap2v = 0 ,xaxs="i",yaxs="i",plot_weights = F)

par(mar=c(4,.1,2,30))
barplot( rev(pvalues) , horiz = T , col= taxa_pal[factor(rev(module_color))],
         xaxs="i",yaxs="i",xlim=c(0,1.2*max(pvalues)),ylab="",xlab="-log10(pvalue)",las=1,names.arg = "")
points( rev(pvalues), seq(1,length(pvalues) )*1.2-.5,pch=21,bg=taxa_pal[factor(rev(module_color))] )
text( rev(pvalues)+1 , seq(1,length(pvalues) )*1.2-.5, adj=0,cex=.8,xpd=T,
      labels = rev(names(pvalues)),xpd=T)
text( rev(pvalues)+1 , seq(1,length(pvalues) )*1.2-.5, adj=0,cex=.8,
      labels = rev(sub(" [(].*","",names(pvalues)) ),col = paste0(taxa_pal[factor(rev(module_color))],90) ,xpd=T)
# text( rep(max(pvalues)+.5,length(pvalues)) , seq(length(pvalues),1 )*1.2-.5,
#       adj=0,cex=.8,labels = names(pvalues), col = taxa_pal[factor(module_color)] )

mypar(2,3,mar=c(2,2,2,6))
set.seed(1)
UMAP_picrust <- uwot::umap(t(named_picrust_data),n_neighbors = 20,
                 metric = "correlation",min_dist = .2,spread = .4,
                 repulsion_strength = .4,negative_sample_rate = 3)
plot(UMAP_picrust,bg=pal[group_annotation[sample_use]],main="UMAP on top functional profile",
     pch=21,xlab="UMAP1",ylab="UMAP2",frame=F,axes=F)
legend(par("usr")[2],par("usr")[4],title.adj = 0,
       legend = levels(factor(group_annotation)),xjust = 0,yjust = 1,
       bty = "n",pch = 21,pt.bg = pal,pt.cex = 1,xpd=T)
```

@@

h <- hclust( as.dist( (1- cor(t(top_TRX_counts)))/2 ), method = "ward.D2")
cuttoff <- 2
gene_module <- cutree(h, h=cuttoff)

layout(matrix(c(1,2,3,4,
              1,2,3,4),
              nrow = 2,ncol = 4,byrow = T),widths = c(1,2,.5,6))
par(mar=c(4,.1,2,.1))
plot( rev(as.dendrogram(h)) , xlim=c(max(h$height),-.2),horiz = T, leaflab = "none",xaxs="i",yaxs="i",axes=T,xlab="height")
# rect.dendrogram(rev(as.dendrogram(h)),h = 2,horiz = TRUE)
abline(v=cuttoff,xpd=F,col="red",lty=2)
points( rep(-.1,length(gene_module)),
        seq(length(gene_module),1,length.out = length(gene_module)),
        col=taxa_pal[factor(gene_module[h$order])],
        pch=16,cex=.5,xpd=T)

image( t(top_TRX_counts[h$order,][nrow(top_TRX_counts):1,order(group_annotation[sample_use])]),
       col = colorRampPalette(c("navy","navy","grey95","firebrick4","firebrick4") )(99),
       breaks = seq(-5,5,length.out = 100),axes=F , xlab="patient groups")
points( seq(0,1,length.out = length(sample_use)),rep(par("usr")[3] - .01,length(sample_use)),
        col=pal[group_annotation[sample_use][ order(group_annotation[sample_use])]],
        pch=16,xpd=T)

source("../enrichment_function.R")
gmt_list <- fgsea::gmtPathways("../../../supplementary_files/c5.bp.v6.2.symbols.gmt.txt")
res_list <- lapply( unique( gene_module[h$order] ),function(x){
  temp <- compute_enrichment(genes = names(gene_module)[gene_module==x],
                            gmt_list = gmt_list,
                            min_terms_pathway = 10,
                            max_terms_pathway = 300,
                            min_overlap = 3,
                            sort_by_pvalue = T)
  temp <- temp[!grepl("REGULATION",rownames(temp)),]
  return(temp)
} )
names(res_list) <- unique( gene_module[h$order] )

pvalues <- unlist(lapply(res_list,function(x){ -log10(x$pvalue) [1:5] }))
pvalues[is.na(pvalues)] <- 0
terms <- unlist(lapply(res_list,function(x){ rownames(x) [1:5] }))
terms[is.na(terms)] <- ""
genes <- unlist(lapply(res_list,function(x){ x$genes [1:5] }))
genes[is.na(genes)] <- ""

pvalues <- setNames(pvalues,paste0(terms," (",genes,")"))
module_color <- unlist(lapply(names(res_list),function(x){ rep(x,5) }))

temp <- factor( rev(gene_module), levels = unique( gene_module[h$order] ))
plot_sankey( data.frame(temp,temp), pal = taxa_pal[as.numeric(levels(temp))] , use_w2 = F , plot_labels = T ,gapv = .01,gap2v = 0 ,xaxs="i",yaxs="i",plot_weights = F)

par(mar=c(4,.1,2,30))

barplot( rev(pvalues) , horiz = T , col= taxa_pal[factor(rev(module_color))],
         xaxs="i",yaxs="i",xlim=c(0,1.2*max(pvalues)),ylab="",xlab="-log10(pvalue)",las=1,names.arg = "")
points( rev(pvalues), seq(1,length(pvalues) )*1.2-.5,pch=21,bg=taxa_pal[factor(rev(module_color))] )
text( rev(pvalues)+1 , seq(1,length(pvalues) )*1.2-.5, adj=0,cex=.8,xpd=T,
      labels = rev(names(pvalues)))
text( rev(pvalues)+1 , seq(1,length(pvalues) )*1.2-.5, adj=0,cex=.8,
      labels = rev(sub(" [(].*","",names(pvalues)) ),col = paste0(taxa_pal[factor(rev(module_color))],90) ,xpd=T)
# text( rep(max(pvalues)+.5,length(pvalues)) , seq(length(pvalues),1 )*1.2-.5,
#       adj=0,cex=.8,labels = names(pvalues), col = taxa_pal[factor(module_color)] )

mypar(2,3,mar=c(2,2,2,6))
set.seed(1)
UMAP_TRX <- uwot::umap(t(top_TRX_counts),n_neighbors = 10,
                 metric = "correlation",min_dist = 0.1,spread = 5,
                 negative_sample_rate = 10)
plot(UMAP_TRX,bg=pal[group_annotation[sample_use]],main="UMAP on top DGE RNAseq",
     pch=21,xlab="UMAP1",ylab="UMAP2",frame=F,axes=F)
legend(par("usr")[2],par("usr")[4],title.adj = 0,
       legend = levels(factor(group_annotation)),xjust = 0,yjust = 1,
       bty = "n",pch = 21,pt.bg = pal,pt.cex = 1,xpd=T)

NESse_list <- read.csv2("../../../results/Hallmark_NESse_list.csv",row.names = 1)
NESse_list <- lapply(unique(sub("[.].*","",colnames(NESse_list))),function(x){
  temp <- NESse_list[ grep( x, colnames(NESse_list)) ]
  colnames(temp) <- sub(".*[.]","",colnames(temp))
  temp[is.na(temp)]<-0
  return(t(temp))
})
pvalue_list <- read.csv2("../../../results/Hallmark_pvalues_list.csv",row.names = 1)
pvalue_list <- lapply(unique(sub("[.].*","",colnames(pvalue_list))),function(x){
  temp <- pvalue_list[ grep( x, colnames(pvalue_list)) ]
  colnames(temp) <- sub(".*[.]","",colnames(temp))
  temp[is.na(temp)]<-0
  return(t(temp))
})
x <- pvalue_list[[1]]
df <- data.frame( pathway=c(sapply(1:ncol(x),x=x,function(i,x){rownames(x)} )),
                  bc=c(sapply(colnames(x),x=x,function(i,x){ rep(i,nrow(x)) }) ),
                  NES=c(NESse_list[[1]]),
                  abs_NES=abs(c(NESse_list[[1]])),
                  pvalue=c(x))
df <- df[df$pvalue!=0,]
df <- df[df$pvalue<0.05,]
df <- df[df$NES!=0,]

df %>% group_by(bc)  %>% top_n(3,NES) -> top_per_bc
df %>% group_by(bc)  %>% top_n(-3,NES) ->  bottom_per_bc

as.data.frame(top_per_bc)

mypar()
x <- NESse_list[[1]] - rowMeans( NESse_list[[1]] )
x [ pvalue_list[[1]] < 0.05 ] <- 0
x <- x[unique(top_per_bc$pathway),]
h <- hclust( as.dist( (1-cor(t(x)))/2 ) , method = "ward.D2" )
image( x[h$order,ncol(x):1],col=colorRampPalette(c("navy","grey95","firebrick"))(91),
       breaks=seq(-5,5,length.out = 92),
       axes=F,border=NA,xlab="",ylab="",line=.4,cex.main=1,font.main=1)
```

@@

empty_plot()
legend(x = par("usr")[c(1)],
         y = par("usr")[c(4)],title.adj = 0,
         legend = rownames(temp),xjust = 0,yjust = 1,
         bty = "n",pch = 22,pt.cex = 2,cex = 1,xpd=T,
          title="top abundant bacteria",
         pt.bg = taxa_pal)
```

@@

## Figure S6

```{r, echo = FALSE, eval = TRUE,fig.asp=1.2}
combined_bc <- rbind( top_per_bc  , bottom_per_bc )

mypar(8,2,mar=c(2,25,1,1))
for(i in sort(levels(combined_bc$bc)) ){
  barplot( c(combined_bc [combined_bc$bc == i,"NES" ])[[1]] , horiz = T,las=1,
           main=i,
           names.arg = as.character(combined_bc$pathway)[combined_bc$bc == i])
}
```
